---
title: "R CODA Club - Creating a descriptives table using gtsummary and other bits and bobs"
author: "Caitlin A. Howlett"
date: "`r Sys.Date()`"
output: pdf_document
---

# Basic descriptives table 

## An example: 


```{r echo = FALSE, include=FALSE, warning=FALSE, message=FALSE}

# ----- load libraries -----

library(tidyverse)
library(gtsummary)
library(dplyr)


# ----- Read in data ------
data <- read.csv("C:/Users/Caitlin/Dropbox/r_projects/R Coda Club/gtsummary/data.csv")


# --- Basic tbl_summary structure ---

# x <-
#   ., %>%
#   tbl_summary(
#     by = ,
#     label = list(),
#     statistic = list(),
#     digits = list(),
#     type = list(),
# missing = "ifany",
# missing_text = "Missing"
# )


# Wherever there are empty cells in the data set, change these cells to "NA" if you want them to count towards missing values

data$gender[data$gender==""] <- NA
table(is.na(data$gender))

# Create a data frame with all of the variables that you want to display in your descriptives table

df1 <-
  data %>%
  select(group, age, gender, education_level, wcst, stroop)


# Build elements of the table 

tab1 <- 
  df1 %>%
  tbl_summary(
    by = group,
    label = list(age ~ "Age in years", gender ~ "Gender", education_level ~ "Education", wcst ~ "WCST score", 
                 stroop ~ "Stroop score"),
    statistic = list(age ~ c("{mean} ({sd})", "{min} - {max}"), gender ~ "{n} ({p}%)", education_level ~ "{n} ({p}%)", 
                     wcst ~ c("{mean} ({sd})", "{min} - {max}"), stroop ~ c("{mean} ({sd})", "{min} - {max}")),
    digits = list(age ~ (c(1, 2, 0, 0)), wcst ~ (c(1, 2, 0, 0)), stroop ~ (c(1, 2, 0, 0))),
    type = list(age ~ "continuous2", gender ~ "categorical", education_level ~ "categorical", wcst ~ "continuous2",
                stroop ~ "continuous2"),
missing = "ifany",
missing_text = "Missing"
) %>%
  bold_labels()

# Modifying aspects of the table 

modify_descriptives_tab1 <-
  tab1 %>%
  modify_header(all_stat_cols() ~ "**{level}**, (n = {n})") %>%
  modify_caption("**Participant Characteristics** (n = {N})") %>%
  modify_footnote(update = starts_with("stat_") ~ "Statistics presented: Mdn (IQR); n(%)")


# Check distribution of continuous values 
hist(data$age)
hist(data$wcst)
hist(data$stroop)

# Replace {mean} ({sd}) with {median} ({p25}, {p75}) for non-normal distributions


# Rename the table for simplicity purposes 
Table1 <- modify_descriptives_tab1

 

```

# Inline code:
 

### Participants

Eight adults in the healthy control group had a mean age of `r inline_text(Table1, variable = age, column = stat_1, level = 'Mean (SD)')`. Participant characteristics are displayed in Table 1. 


```{r Table 1, echo=FALSE, results='asis', message=FALSE, warning=FALSE}

Table1

```


# Stacking tables together 

```{r echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}

# Print out of how many different types of conditions (and spelling variations) there are

table(data$diagnosis, useNA = "ifany")

# Turning different spelling/capitalisation variations of responses into a TRUE or FALSE conditional statement, so that we can count the number of occurrences of a particular condition 

# Fibromyalgia 

data <-
  data %>%
  mutate(
    diagnosis_fibro = 
      case_when(
        diagnosis == "fibro" ~ 1,
        diagnosis == "fibromyyalgia" ~ 1,
        diagnosis == "Fibromyalgia" ~ 1,
        TRUE ~ 0
      )
  )

with(data,
     table(
       diagnosis, 
       diagnosis_fibro, 
       useNA = "ifany"
     ))


# CRPS

data <-
  data %>%
  mutate(
    diagnosis_crps = 
      case_when(
        diagnosis == "CRPS" ~ 1,
        TRUE ~ 0
      )
  )

with(data,
     table(
       diagnosis, 
       diagnosis_crps, 
       useNA = "ifany"
     ))


# Endometriosis

data <-
  data %>%
  mutate(
    diagnosis_endo = 
      case_when(
        diagnosis == "endometriosis" ~ 1,
        TRUE ~ 0
      )
  )

with(data,
     table(
       diagnosis, 
       diagnosis_endo, 
       useNA = "ifany"
     ))



# Migraine 

data <-
  data %>%
  mutate(
    diagnosis_mig = 
      case_when(
        diagnosis == "migraine" ~ 1,
        TRUE ~ 0
      )
  )

with(data,
     table(
       diagnosis, 
       diagnosis_mig, 
       useNA = "ifany"
     ))




data <- 
  data %>%
  select(id, group, age, gender, education_level, wcst, stroop, diagnosis_fibro, diagnosis_crps, diagnosis_mig, diagnosis_endo)


# Rename variables ----
data <- rename(data, `Fibromyalgia` = diagnosis_fibro, `Endometriosis` = diagnosis_endo, `Migraine` = diagnosis_mig, `CRPS` = diagnosis_crps)

```




```{r echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}


# Wherever there are empty cells in the data set, change these cells to "NA" if you want them to count towards missing values
data$gender[data$gender==""] <- NA
table(is.na(data$gender))

# Create a data frame with all of the variables that you want to display in your descriptives table

df2 <- 
  data %>%
  select(id, group, age, gender, education_level, wcst, stroop, Fibromyalgia, CRPS, Migraine, Endometriosis)


# Create a long data set so that we can count the number of occurrences of a condition for each participant 

df3 <- 
  df2 %>%
  pivot_longer(
    c(-id, -group, -age, -gender, -education_level, -wcst, -stroop),
    names_to = "pain_diagnosis",
    values_to = "count") %>%
  mutate(
    paindiag_str = ifelse(count == 1, pain_diagnosis, "")
  )


# Create two separate data frames 
    
descrip_dat2 <-
  df2 %>%
  select(group, age, gender, education_level, wcst, stroop)

descrip_dat3 <-
  df3 %>%
  select(group, paindiag_str)


# Build the first table (containing our demographic and outcome variables)

tab2 <- 
  descrip_dat2 %>%
  tbl_summary(
    by = group,
    label = list(age ~ "Age in years", gender ~ "Gender", education_level ~ "Education", wcst ~ "WCST score", stroop ~ "Stroop score"),
    statistic = list(age ~ c("{mean} ({sd})", "{min} - {max}"), gender ~ "{n} ({p})", education_level ~ "{n} ({p})", 
                     wcst ~ c("{mean} ({sd})", "{min} - {max}"), stroop ~ c("{mean} ({sd})", "{min} - {max}")),
    digits = list(age ~ (c(1, 2, 0, 0)), wcst ~ (c(1, 2, 0, 0)), stroop ~ c(1, 2, 0, 0)),
    type = list(age ~ "continuous2", gender ~ "categorical", education_level ~ "categorical", wcst ~ "continuous2",
                stroop ~ "continuous2"),
missing = "ifany",
missing_text = "Missing"
) %>%
  bold_labels()


# Modify elements of the first descriptives table 

modify_descriptives_table2 <-
  tab2 %>%
  modify_header(all_stat_cols() ~ "**{level}**, (n = {n})") %>%
  modify_caption("**Participant Characteristics** (n = {N})")


tab2 <- modify_descriptives_table2


# Build the second table (containing counts of pain diagnoses)

n_hc <- 8 
n_pp <- 8

tab3 <-
  descrip_dat3 %>%
  tbl_summary(
    by = group,
    label = list(paindiag_str ~ "Pain diagnosis"),
    statistic = list(paindiag_str ~ "{n}"),
    type = list(paindiag_str ~ "categorical"),
    missing = "ifany",
    missing_text = "Missing",
    include = everything()
  ) %>%
  bold_labels() 
  

# Modify elements of the second descriptives table

  tab3$table_body <-
  tab3$table_body %>%
  # get rid of empty levels 
  dplyr::filter(label != "") %>%
  # add proportions based on column sample size
  mutate(
    stat_1 =
      ifelse(is.na(stat_1), NA, paste0(stat_1, " (", round(100 * as.numeric(stat_1) / n_hc, 0), ")")),
    stat_2 =
    ifelse(is.na(stat_2), NA, paste0(stat_2, " (", round(100 * as.numeric(stat_2) / n_pp, 0), ")"))
  )
  

Table2 <-
    tbl_stack(list(
      modify_descriptives_table2,
      tab3
    )) %>%
    as_kable(.)
  

```



```{r Table1, echo=FALSE, results='asis'}

Table2
```



# Other useful packages:

- ggstatsplot (presents linear models as visual plots)
- jtools (similar to ggstatsplot but overlays two models onto a single visual plot)
- report (automated reporting of results from statistical tests)


```{r echo = FALSE, include=FALSE, warning=FALSE, message=FALSE}

# Convert gender to a factor

with(data, table(gender, useNA = "ifany"))
data$gender <- factor(data$gender)
data$gender <- relevel(data$gender, ref = "Male") 

class(data$gender)

```
```{r echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}

# Convert education level to a factor

with(data, table(education_level, useNA = "ifany"))
data$education_level <- factor(data$education_level)
data$education_level <- relevel(data$education_level, ref = "High school") 

class(data$education_level)

```
```{r multiple linear regression model, echo=FALSE, message=FALSE, warning=FALSE}

library(performance)

wcst_model <- lm(wcst ~ group + age + education_level, data = data)

stroop_model <- lm(stroop ~ group + age + education_level, data = data)

summary(wcst_model)
summary(stroop_model)

car::Anova(wcst_model)
car::Anova(stroop_model)

# Check model assumptions 

check_model(wcst_model)
check_model(stroop_model)


```


```{r echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}

Table3 <- lm(wcst ~ group + age + education_level + gender, data = data) %>%
  tbl_regression() %>%
  add_global_p()


```

```{r echo=FALSE, warning=FALSE, message=FALSE}

Table3

```

```{r Visualising and reporting final WCST model, echo=FALSE, warning=FALSE, message=FALSE, include=TRUE}


# regression plots

library(ggstatsplot)
ggcoefstats(wcst_model)

# how to change colour palette of plots to make them accessible for people who are have vision deficiencies

ggcoefstats(wcst_model, package = "RColorBrewer", palette = "Dark2")
ggcoefstats(wcst_model, package = "colorBlindness", palette = "paletteMartin")
ggcoefstats(wcst_model, package = "colorblindr", palette = "OkabeIto")


library(jtools)
plot_coefs(wcst_model, stroop_model, omit.coefs = NULL, model.names = c("WCST model", "Stroop model"),
                                                                                       ci_level = 0.95)
# mathematical equations 

library(equatiomatic)
extract_eq(wcst_model, use_coefs = TRUE, wrap = TRUE, terms_per_line = 2)
extract_eq(stroop_model, use_coefs = TRUE, wrap = TRUE, terms_per_line = 2)


# print summaries of results for academic reporting 

library(report)
report(wcst_model)
report(stroop_model)

```





# Useful resources for gtsummary: 

- gtsummary paper: https://journal.r-project.org/archive/2021/RJ-2021-053/RJ-2021-053.pdf

- gtsummary cheatsheet: https://www.rstudio.com/resources/cheatsheets/

- R Ladies STL recently held a workshop with the package developer of gtsummary (Daniel Sjoberg) - the workshop was posted on the R Ladies STL Youtube channel, which can be found here: https://www.youtube.com/channel/UCQ7DcXWJm3es06U7lGmBDkw 




# Citing R packages:

citation("tidyverse")

citation("gtsummary")

citation("report")

citation("equatiomatic")

citation("jtools")

citation("ggstatsplot")


